\chapter{Date and Time}
\label{sec:datetime}

In this Chapter we take a little break and concentrate on a topic that it is not usually covered in a \texttt{python} review for beginners. However given its importance for financial computation the next paragraphs will be devoted to a close look up on the \texttt{datetime} module~\cite{bib:datetime}, whose usage will help in manipulating dates.

\section{Dates}\label{dates}

As said dates are not usually included in a standard \texttt{python} tutorial, however since they 
are pretty essential for finance we are going to cover this topic in some detail. 
In \texttt{python} the date utilities mainly lives in the \texttt{datetime} module. Briefly we are 
also going to show \texttt{relativedelta} from the \texttt{dateutil} module, which allows to 
add/subtract days/months/years to dates, in other words to make operations on them.

In this first example today's date is defined and with \texttt{relativedelta} two more dates are created 
adding two months and subtracting three days to the first one.

\begin{ipython}
from datetime import date, datetime
from dateutil.relativedelta import relativedelta

date1 = date.today()
print(date1)

date2 = date.today() + relativedelta(months=2)
print(date2)

date3 = date.today() - relativedelta(days=3)
print(date3)
\end{ipython}
\begin{ioutput}
2020-08-03
2020-10-03
2020-07-31
\end{ioutput}

There is another way of computing a new date: a one day delta is stored in a variable 
and today's date is moved by three days multiplying the defined delta by three.

\begin{ipython}
one_day = relativedelta(days=1)
date.today() - 3 * one_day
\end{ipython}
\begin{ioutput}
datetime.date(2020, 7, 31)
\end{ioutput}

Next, given two dates their difference is computed and expressed in number of days.

\begin{ipython}
date1 = date(2019, 7, 2)
date2 = date(2019, 8, 16)
(date2 - date1).days
\end{ipython}
\begin{ioutput}
45
\end{ioutput}

Dates can be converted to and from strings and a large variety of formats can be specified in this conversions. 
The format is determined by a string in which each character starting with \% represent an element 
of the date, e.g. \%Y year, \%d day, \%s seconds\ldots

Below dates to string conversion:

\begin{ipython}
date1 = date(2019, 7, 2)
date1.strftime("%Y-%b-%d (%a)") # dates can formatted in many ways
                                # check the docs for more details
\end{ipython}
\begin{ioutput}
'2019-Jul-02 (Tue)'
\end{ioutput}

And here, a string is converted to \texttt{datetime} object:

\begin{ipython}
# a string can be converted to dates too
datetime.strptime('25 Aug 2019', "%d %b %Y").date()
\end{ipython}
\begin{ioutput}
datetime.date(2019, 8, 25)
\end{ioutput}

Finally a last example showing how to get the week-day from a date:

\begin{ipython}
date1.weekday() # 0 = Monday, ..., 6 = Sunday
\end{ipython}
\begin{ioutput}
1
\end{ioutput}

\section{Exercises}
\input{datetime_ex_text}

\begin{thebibliography}{9}
\bibitem{bib:datetime} \href{https://realpython.com/python-datetime/}{\emph{Using Python datetime to Work With Dates and Times}}, Real Python [Online]
\end{thebibliography}