\chapter{Var and Credit Risk}\label{lesson-13}


\hypertarget{exercise-9.1}{%
	\subsubsection{Exercise 9.1}\label{exercise-9.1}}

Given the historical series of two stock prices in the file
\(\href{https://repl.it/@MatteoSani/support9}{\textrm{historical\_data.py}}\)
compute the 5-day 95\% VaR for a portfolio consisting of 100 shares of
stock 1 and 50 shares of stock 2 (assume that last price of the series
is today's price).

\hypertarget{exercise-9.2}{%
	\subsubsection{Exercise 9.2}\label{exercise-9.2}}

Imagine a position consisting of 500000 EUR investemnt in FCA shares and
a 750000 investiment in Apple shares. Assume that the daily volatilites
of the two assets are 2.5\% and 0.7\% and that their correlation
coefficient is 0.4.

What is tha 10-day 97.5\% VaR for the portfolio ?

\hypertarget{exercise-9.3}{%
	\subsubsection{Exercise 9.3}\label{exercise-9.3}}
Find today's price of a 4-years bond which pays semiannual coupons indexed with the LIBOR curve defined in \(\href{https://repl.it/@MatteoSani/support9}{\textrm{curve\_data.py}}\). The face value of the bond is 100000 EUR.

\hypertarget{exercise-9.4}{%
	\subsubsection{Exercise 9.4}\label{exercise-9.4}}

Given an arbitrary number of students and some project label
write a python program which assigns randomly a project to each student
(repetition of projects are allowed).


\begin{Exercise}
Update \verb!finmarkets! module with \verb!CreditCurve! and \verb!CreditDefaultSwap! classes.
\end{Exercise}
\begin{Answer}
\end{Answer}
