\chapter{Modeling Correlation between Risks}

In credit derivative valuation and credit risk management, one of the
fundamentally important issues is the estimate of default
probabilities and their correlations. For this, generally speaking,
there are two ways: using historical default data or using mathematical
models.

Historical default data has played an important role in the estimation
of default probabilities. However, because default events are rare,
there is very limited default data available. Moreover, historical data
reflects the historical default pattern only and it may not be a proper
indicator of the future. This makes the estimate of default
probabilities from historical data difficult and inexact. To use this
same data to estimate default correlations is even more difficult and
more inexact.

The market trend now is towards more and more to the use of mathematical
models that don't rely on historical default data. In
Chapter~\ref{credit_default_swaps} we have seen how it is possible to derive default 
probabilities from market data,
here we will see how the copula can be used to model their correlations. 

\section{One Factor Gaussian Copula Model}\label{standard-market-model}
While there are several types of copula function models, the first
introduced was the \emph{one-factor Gaussian copula model}. This model has,
above all, the advantage that can be solved semi-analytically.

Consider a portfolio of \(N\) bonds and assume that the marginal
probabilities of default are known for each issuer. Define:

\begin{itemize}
	\tightlist
	\item
	\(t_i\), the time of default of the \(i^{th}\) company:
	\item
	\(Q_i(t)\), the cumulative probability that company \(i\) will default
	before time \(t\); that is, the probability that \(t_i \le t\);
	\item
	\(S_i(t) = 1 - Q_i(t)\), the probability that company \(i\) will
	survive beyond time \(t\); that is, the probability that \(t_i > t\).
\end{itemize}

To generate a one-factor model for the \(t_i\) we define random
variables \(X_i\) \((1\le i \le N)\)
\begin{equation}
X_i = a_i M + \sqrt{1-a_i^2}Z_i,\qquad i = 1, 2,\ldots, n
\label{eq:normalized_var}
\end{equation}
where \(M\) and the \(Z_i\) are independent zero-mean unit-variance  distributions (hence $X_i$ are also distributed with zero-mean and unit standard-deviation) and \(-1 \le a_i \lt 1\).

The previous equation defines a correlation structure between the
\(X_i\) which are dependent on a single common factor \(M\). The $Z_i$ term is usually 
called the idiosincratic component of default. 
The correlation between \(X_i\) and \(X_j\) is

\[
\mathrm{Corr}(X_i, X_j) = \cfrac{\mathbb{E}[(X_i-\mu_i)(X_j-\mu_j)]}{\sigma_{X_i}\sigma_{X_j}} =\mathbb{E}[X_i X_j] = a_i a_j \mathbb{E}[M^2] = a_i a_j
\]
where we just exploit the definition of $X_i$ and its properties.

Assume that the $i^{th}$ company has defaulted by the time $t_i$ if $X_i$ is below a threshold value $\bar{x}_i(t_i)$.
If $F$ is the cumulative distribution function of the $X_i$,
with a percentile to percentile transformation we can map the \(X_i\) to the \(t_i\), so that $Q_i(t_i) = \mathbb{P}(X_i\le x)=F(x)$.
Therefore the point \(X_i = x\) is transformed to \(t_i = t\) where
\(x = F_i^{–1}[Q_i(t)]\).

Let's note the, \emph{conditional} on $M$, the $N$ default events are independent. So we can write
\begin{equation*}
\begin{split}
Q_i(t_i|M) &= \mathbb{P}(X_i\le x|M) = \mathbb{P}(a_i M + \sqrt{1-a_i^2}Z_i\le x) \\
&= \mathbb{P}\left(Z_i\le \cfrac{x-a_i M}{\sqrt{1-a_i^2}}\right)
=H_i\left(\cfrac{F^{-1}[Q(t_i)]-a_i M}{\sqrt{1-a_i^2}}\right)
\end{split}
\end{equation*}
where $H_i$ is the cumulative distribution function of the $Z_i$.

Although in principle any distribution could be used for \(M\)'s and the
\(Z\)'s (provided they have zero mean and unit variance), one common
choice is to let them be standard normal distributions (resulting in a
Gaussian copula).
So we can rewrite the previous equation as

\[
Q_i(t_i|M) = \Phi\left(\cfrac{\Phi^{-1}[Q(t_i)]-a_i M}{\sqrt{1-a_i^2}}\right)
\]
where $\Phi$ denotes the cumulative distribution function of the standard normal distribution.

If we call $\mathcal{C}(t_1,\ldots,t_N)$ the joint distribution of the default times of the $N$ bonds  in the portfolio then

\[
\mathcal{C}(t_1,\ldots,t_N)=\Phi_{A}(\Phi^{-1}(Q_1(t_1),\ldots,\Phi^{-1}(Q_N(t_N)))
\]
where $A$ is the correlation matrix of the default probabilities, is the one factor Gaussian copula model (one factor because the is only a random variable, $M$, which determines the correlation between $X_i$).

Clearly different choices of distributions result in different copula models, and in different natures of the default dependence. For example, copulas where the \(M\)'s have heavy tails generate models where there is a
greater likelihood of a clustering of early defaults for several
companies.

\subsection{Standard Market Model}\label{standard-market-model}

Assume the following two assumptions are made:

\begin{itemize}
	\tightlist
	\item
	all the companies have the same default intensity (hazard rates), i.e, \(\lambda_i = \lambda\) (which means they all have the same default probabilities);
	\item
	the pairwise default correlations are the same, i.e \(a_i = a\); in other word the contribution of the market
	component $M$ is the same for all the companies and the correlation between any two companies is constant, \(\rho = a^2\).
\end{itemize}

Under these assumptions, given the market situation \(M = m\), all the
companies have the same cumulative default probability
\(DP_{t|M}=Q_i(t_i|M)=\mathbb{P}(X_i < x|M)\). 
Moreover, for a given value of the
market component \(M\), the defaults are mutually independent for all
the underlying companies. 

Letting \(N_{t|m}\) be the total defaults that
have occurred by time \(t\) conditional on the market condition
\(M = m\), then \(N_{t|m}\) follows a binomial distribution, and

\[DP(N_{t|m} = j) = \cfrac{n!}{j!(n-j)!}DP^j_{t|m}(1-DP_{t|m})^{n-j},\qquad  j=0, 1, 2,\ldots,n\]
The probability that there will be exactly \(j\) defaults by time \(t\)
is
\begin{equation}
DP(N_{t} = j) = \int_{-\infty}^{\infty}{DP(N_{t|m} = j)f_M(m)dm}
\label{eq:gaussian_quadrature}
\end{equation}
where \(f_M(m)\) is the probability density function (PDF) of the random
variable \(M\).

With the assumption that have been made the one factor model is also called \emph{Market Standard Model}.


%\subsubsection{Gaussian Quadrature}\label{gaussian-quadrature}
%The integral in Eq.~\ref{eq:gaussian_quadrature} can be quite
%complicated depending on the distribution of \(f_M\).
%
%The Gaussian Quadrature is a technique that allows to approximate that
%integral with a discrete weighted sum with weights determined by the
%function \(f_M\). Assuming \(f\) is Gaussian we can write
%
%\[\int_{-\infty}^{+\infty}\cfrac{1}{\sqrt{2\pi}}e^{-F^{2}/2}g(F)dF\approx\sum_{k=1}^{k=N}w_k g(F_k)\]
%
%As \(N\) increases the accuracy of the approximation increases, but usually
%\(N=60\) is sufficient.
%
%\href{https://drive.google.com/file/d/1Ic20cgVx4dpDG4W_pIHR4EXq9ens1G-H/view?usp=sharing}{Here} a dedicated class, \(\tt{GaussianQuadrature}\) is
%available. Through the method \(\tt{M}\) returns the appropriate list of
%weights \(w_k\) and values \(F_k\) to compute the intergral (it is
%necessary to download also \href{https://drive.google.com/file/d/1zpQ0ubbEzniJb9usMWeeMCc0DSc2MGsC/view?usp=sharing}{this file} with all the values).
%
%\begin{tcolorbox}[breakable, size=fbox, boxrule=1pt, pad at break*=1mm,colback=cellbackground, colframe=cellborder]
%\begin{Verbatim}[commandchars=\\\{\}]
%\PY{k+kn}{from} \PY{n+nn}{finmarkets} \PY{k}{import} \PY{n}{GaussianQuadrature}
%	
%\PY{n}{gq} \PY{o}{=} \PY{n}{GaussianQuadrature}\PY{p}{(}\PY{p}{)}
%\PY{n+nb}{print} \PY{p}{(}\PY{l+s+s2}{\PYZdq{}}\PY{l+s+s2}{values: }\PY{l+s+s2}{\PYZdq{}}\PY{p}{,} \PY{n}{gq}\PY{o}{.}\PY{n}{M}\PY{p}{(}\PY{l+m+mi}{60}\PY{p}{)}\PY{p}{[}\PY{l+m+mi}{0}\PY{p}{]}\PY{p}{)}
%\PY{n+nb}{print} \PY{p}{(}\PY{l+s+s2}{\PYZdq{}}\PY{l+s+s2}{weights: }\PY{l+s+s2}{\PYZdq{}}\PY{p}{,} \PY{n}{gq}\PY{o}{.}\PY{n}{M}\PY{p}{(}\PY{l+m+mi}{60}\PY{p}{)}\PY{p}{[}\PY{l+m+mi}{1}\PY{p}{]}\PY{p}{)}
%
%values:  [14.36715008, 13.46459111, 12.71717121, 12.04990464, 11.43418345,
%10.85527694, 10.30435118, 9.775583711, 9.264879565, 8.769218063, 8.286287662,
%7.814266859, 7.351685308, 6.897331785, 6.450190946, 6.009398508, 5.574208647,
%5.143969727, 4.718105868, 4.296102681, 3.877496046, 3.461863141, 3.048815156,
%2.637991292, 2.229053735, 1.821683385, 1.415576157, 1.010439725, 0.605990602,
%0.201951448, -0.201951448, -0.605990602, -1.010439725, -1.415576157,
%-1.821683385, -2.229053735, -2.637991292, -3.048815156, -3.461863141,
%-3.877496046, -4.296102681, -4.718105868, -5.143969727, -5.574208647,
%-6.009398508, -6.450190946, -6.897331785, -7.351685308, -7.814266859,
%-8.286287662, -8.769218063, -9.264879565, -9.775583711, -10.30435118,
%-10.85527694, -11.43418345, -12.04990464, -12.71717121, -13.46459111,
%-14.36715008]
%weights:  [6.26018e-46, 1.37648e-40, 2.12791e-36, 7.51816e-33, 9.67908e-30,
%5.8078e-27, 1.88764e-24, 3.67432e-22, 4.6002e-20, 3.90602e-18, 2.34277e-16,
%1.02492e-14, 3.35604e-13, 8.40054e-12, 1.63579e-10, 2.51449e-09, 3.08925e-08,
%3.06553e-07, 2.47921e-06, 1.64672e-05, 9.04268e-05, 0.000412859, 0.001574836,
%0.005039442, 0.013575119, 0.030871873, 0.059408466, 0.096914318, 0.134203158,
%0.157890214, 0.157890214, 0.134203158, 0.096914318, 0.059408466, 0.030871873,
%0.013575119, 0.005039442, 0.001574836, 0.000412859, 9.04268e-05, 1.64672e-05,
%2.47921e-06, 3.06553e-07, 3.08925e-08, 2.51449e-09, 1.63579e-10, 8.40054e-12,
%3.35604e-13, 1.02492e-14, 2.34277e-16, 3.90602e-18, 4.6002e-20, 3.67432e-22,
%1.88764e-24, 5.8078e-27, 9.67908e-30, 7.51816e-33, 2.12791e-36, 1.37648e-40,
%6.26018e-46]
%\end{Verbatim}
%\end{tcolorbox}


\subsection{Extensions of the One Factor Copula Model}
Many other one-factor model have been tried: Student t copula, Clayton copula and many others. In general we can define a new model by simply choosing particular functions for $M$ and $Z_i$ in Eq.~\ref{eq:normalized_var} provided they are with mean zero and standard deviation 1. 

If instead of the single factor $M$ there are two or more, Eq.~\ref{eq:normalized_var} would become

\[
X_i = a_1 M_1 + a_2 M_2 + \sqrt{1 - a_1^2 - a^2_2}Z_i
\]
and similarly
\[
Q(t|M_1, M_2) = \Phi\left(\cfrac{\Phi^{-1}[Q(t)]-a_1 M_1 - a_2 M_2}{\sqrt{1 - a_1^2 - a^2_2}}\right)
\]
This kind of models are proportionally slower with the increase of the number of factors.

Later in this Chapter we will also see that the copula can be implied from the market quotes of CDOs.

\section{Basket Default Swaps}\label{basket-default-swaps}

A basket default swap is a credit derivative on a portfolio of reference
entities. The simplest basket default swaps are first-to-default,
second-to-default, or nth-to-default swaps. 

This kind of contracts are very similar to normal CDS except for the protection they offer.
With respect to a basket of reference entities, a first-to-default swap provides insurance for only the first default, a second-to-default swap provides insurance
for only the second default, and a nth-to-default swap provides insurance for only the nth default. 

For example, in the last case, the
seller does not make a payment to the protection buyer for
the first n-1 defaulted reference entities, and makes a payment only for the
$n^{th}$ defaulted reference entity. Once there has been this payment the swap terminates.

\section{Correlated Defaults}\label{correlated-defaults}
The cost of protection in nth-to-default CDS is obviously dependent
on default correlations. 

Suppose that a basket of 100 reference entities is used to define a 5-year nth-to-default CDS and that each reference entity has a probability of 2\% of defaulting during the next
5 years. When the default correlation between the reference entities is zero the binomial distribution (see Appendix~\ref{binomial-distribution}) 
shows that the probability of one or more defaults during 5 years is 86.74\% and the probability of ten or more defaults is 0.0034\%.

\begin{tcolorbox}[breakable, size=fbox, boxrule=1pt, pad at break*=1mm,colback=cellbackground, colframe=cellborder]
\begin{Verbatim}[commandchars=\\\{\}]
\PY{k+kn}{from} \PY{n+nn}{scipy}\PY{n+nn}{.}\PY{n+nn}{stats} \PY{k}{import} \PY{n}{binom}
	
\PY{n}{b} \PY{o}{=} \PY{n}{binom}\PY{p}{(}\PY{l+m+mi}{100}\PY{p}{,} \PY{l+m+mf}{0.02}\PY{p}{)}
\PY{n+nb}{print}\PY{p}{(}\PY{l+s+s2}{\PYZdq{}}\PY{l+s+s2}{P(\PYZgt{}=1) : }\PY{l+s+si}{\PYZob{}\PYZcb{}}\PY{l+s+s2}{\PYZdq{}}\PY{o}{.}\PY{n}{format}\PY{p}{(}\PY{l+m+mi}{1} \PY{o}{\PYZhy{}} \PY{n}{b}\PY{o}{.}\PY{n}{cdf}\PY{p}{(}\PY{l+m+mi}{0}\PY{p}{)}\PY{p}{)}\PY{p}{)}
\PY{n+nb}{print}\PY{p}{(}\PY{l+s+s2}{\PYZdq{}}\PY{l+s+s2}{P(\PYZgt{}=10): }\PY{l+s+si}{\PYZob{}\PYZcb{}}\PY{l+s+s2}{\PYZdq{}}\PY{o}{.}\PY{n}{format}\PY{p}{(}\PY{l+m+mi}{1} \PY{o}{\PYZhy{}} \PY{n}{b}\PY{o}{.}\PY{n}{cdf}\PY{p}{(}\PY{l+m+mi}{9}\PY{p}{)}\PY{p}{)}\PY{p}{)}
	
P(>=1) : 0.8673804441052471
P(>=10): 3.441680604299169e-05
\end{Verbatim}
\end{tcolorbox}

A first-to-default is therefore quite valuable whereas a tenth-to-default CDS is worth almost nothing.

As the default correlation increases the probability of one or more defaults declines and the probability of ten or more defaults increases. In the limit where the default correlation is perfect the probability of one or more defaults equals the probability of ten or more defaults and it is 2\%. This is because in this extreme situation the reference entities are essentially the same : either they all default (with 2\% probability) or none of them default (with 98\% probability).

\subsection{Calculating nth-to-default probability}

\subsubsection{Independent Defaults}\label{independent-defaults}

If the default times of the names of a basket are independent,
first-to-default, nth-to-default, all-to-default probabilities can be
calculated through multiplication and integration of the default
probability curves of the basket components.

As an example, consider the second-to-default probability of a 4-name
basket. Let \(\tau_i\) be the default time of name \(i\) and \(F_i(t)\)
its distribution. Then the probability that name 1 defaults second in
the basket before time \(t\) is:

\begin{equation}
\begin{split}
&\mathbb{P}((\tau_2\lt\tau_1)\cap (\tau_1\lt t)\cap (\tau_1\lt\tau_3)\cap (\tau_1\lt\tau_4)) +\\
&\mathbb{P}((\tau_3\lt\tau_1)\cap (\tau_1\lt t)\cap (\tau_1\lt\tau_2)\cap (\tau_1\lt\tau_4)) =\\
&\int_0^t{F_2 (s)\cdot (1-F_3 (s)) \cdot (1-F_4 (s))~dF_1(s)} +  \int_0^t{F_3 (s)\cdot (1-F_2 (s)) \cdot (1-F_4 (s))~dF_1(s)}
\end{split}
\label{eq:indep_default}
\end{equation}

Suppose the default probabilities of three companies, $A$, $B$ and $C$ are
given as in the following table (in each interval are linear):

\begin{center}
	\begin{tabular}{|c|c|c|c|}
		time in years & A & B & C \\
		\hline
		0 & 0 & 0 & 0 \\
		1 & 0.022032 & 0.0317 & 0.035 \\
		2 & 0.046242 & 0.0655 & 0.075 \\
		3 & 0.07266 & 0.1022 & 0.121 \\
		4 & 0.101233 & 0.142 & 0.153 \\
		5 & 0.131885 & 0.1752 & 0.205 \\
	\end{tabular}
\end{center}
and suppose that the default events of the three companies are
independent. The integral in Eq.~\ref{eq:indep_default} can be solved by substitution:

\[ \int_{x_0}^{x_1}{(1-F_B(x))(1-F_C(x))dF_A(x)}\]

Setting \(t=m_A x + q_A\) it becomes:

\[ \int_{m_A x_0 + q_A}^{m_A x_1 + q_A}{(1-F_B(x(t)))(1-F_C(x(t)))dt}~~~~~~\Big(\textrm{with}~x(t) = \cfrac{t -q_A}{m_A}\Big) \]
and similarly for company $B$ and $C$.

To convert it into \texttt{python} we can use \texttt{scipy.integrate.quad} to
perform the integral and \texttt{numpy.interp} to determine the
intermediate default probabilities.

\begin{tcolorbox}[breakable, size=fbox, boxrule=1pt, pad at break*=1mm,colback=cellbackground, colframe=cellborder]
\begin{Verbatim}[commandchars=\\\{\}]
\PY{k+kn}{from} \PY{n+nn}{scipy}\PY{n+nn}{.}\PY{n+nn}{integrate} \PY{k}{import} \PY{n}{quad}
\PY{k+kn}{from} \PY{n+nn}{numpy} \PY{k}{import} \PY{n}{interp}
	
\PY{n}{default\PYZus{}rates} \PY{o}{=} \PY{p}{\PYZob{}}\PY{l+s+s2}{\PYZdq{}}\PY{l+s+s2}{A}\PY{l+s+s2}{\PYZdq{}}\PY{p}{:}\PY{p}{(}\PY{l+m+mi}{0}\PY{p}{,} \PY{l+m+mf}{0.022032}\PY{p}{,} \PY{l+m+mf}{0.046242}\PY{p}{,} \PY{l+m+mf}{0.07266}\PY{p}{,} \PY{l+m+mf}{0.101233}\PY{p}{,} \PY{l+m+mf}{0.131885}\PY{p}{)}\PY{p}{,}
                \PY{l+s+s2}{\PYZdq{}}\PY{l+s+s2}{B}\PY{l+s+s2}{\PYZdq{}}\PY{p}{:}\PY{p}{(}\PY{l+m+mi}{0}\PY{p}{,} \PY{l+m+mf}{0.0317}\PY{p}{,} \PY{l+m+mf}{0.0655}\PY{p}{,} \PY{l+m+mf}{0.1022}\PY{p}{,} \PY{l+m+mf}{0.142}\PY{p}{,} \PY{l+m+mf}{0.1752}\PY{p}{)}\PY{p}{,}
                \PY{l+s+s2}{\PYZdq{}}\PY{l+s+s2}{C}\PY{l+s+s2}{\PYZdq{}}\PY{p}{:}\PY{p}{(}\PY{l+m+mi}{0}\PY{p}{,} \PY{l+m+mf}{0.035}\PY{p}{,} \PY{l+m+mf}{0.075}\PY{p}{,} \PY{l+m+mf}{0.121}\PY{p}{,} \PY{l+m+mf}{0.153}\PY{p}{,} \PY{l+m+mf}{0.205}\PY{p}{)}\PY{p}{\PYZcb{}}
	
\PY{k}{def} \PY{n+nf}{func}\PY{p}{(}\PY{n}{x}\PY{p}{,} \PY{n}{default}\PY{p}{,} \PY{n}{companies}\PY{p}{,} \PY{n}{t}\PY{p}{)}\PY{p}{:}
    \PY{n}{m} \PY{o}{=} \PY{n}{default}\PY{p}{[}\PY{n}{companies}\PY{p}{[}\PY{l+m+mi}{0}\PY{p}{]}\PY{p}{]}\PY{p}{[}\PY{n}{t}\PY{p}{]} \PY{o}{\PYZhy{}} \PY{n}{default}\PY{p}{[}\PY{n}{companies}\PY{p}{[}\PY{l+m+mi}{0}\PY{p}{]}\PY{p}{]}\PY{p}{[}\PY{n}{t}\PY{o}{\PYZhy{}}\PY{l+m+mi}{1}\PY{p}{]}
    \PY{n}{q} \PY{o}{=} \PY{n}{default}\PY{p}{[}\PY{n}{companies}\PY{p}{[}\PY{l+m+mi}{0}\PY{p}{]}\PY{p}{]}\PY{p}{[}\PY{n}{t}\PY{o}{\PYZhy{}}\PY{l+m+mi}{1}\PY{p}{]} \PY{o}{\PYZhy{}} \PY{n}{m} \PY{o}{*} \PY{p}{(}\PY{n}{t}\PY{o}{\PYZhy{}}\PY{l+m+mi}{1}\PY{p}{)}
    \PY{n}{t} \PY{o}{=} \PY{p}{(}\PY{n}{x}\PY{o}{\PYZhy{}}\PY{n}{q}\PY{p}{)}\PY{o}{/}\PY{n}{m}
    \PY{n}{F2} \PY{o}{=} \PY{l+m+mi}{1} \PY{o}{\PYZhy{}} \PY{n}{interp}\PY{p}{(}\PY{n}{t}\PY{p}{,} \PY{n+nb}{range}\PY{p}{(}\PY{n+nb}{len}\PY{p}{(}\PY{n}{default}\PY{p}{[}\PY{n}{companies}\PY{p}{[}\PY{l+m+mi}{1}\PY{p}{]}\PY{p}{]}\PY{p}{)}\PY{p}{)}\PY{p}{,} \PY{n}{default}\PY{p}{[}\PY{n}{companies}\PY{p}{[}\PY{l+m+mi}{1}\PY{p}{]}\PY{p}{]}\PY{p}{)}
    \PY{n}{F3} \PY{o}{=} \PY{l+m+mi}{1} \PY{o}{\PYZhy{}} \PY{n}{interp}\PY{p}{(}\PY{n}{t}\PY{p}{,} \PY{n+nb}{range}\PY{p}{(}\PY{n+nb}{len}\PY{p}{(}\PY{n}{default}\PY{p}{[}\PY{n}{companies}\PY{p}{[}\PY{l+m+mi}{2}\PY{p}{]}\PY{p}{]}\PY{p}{)}\PY{p}{)}\PY{p}{,} \PY{n}{default}\PY{p}{[}\PY{n}{companies}\PY{p}{[}\PY{l+m+mi}{2}\PY{p}{]}\PY{p}{]}\PY{p}{)}
    \PY{k}{return} \PY{n}{F2}\PY{o}{*}\PY{n}{F3}
	
\PY{k}{def} \PY{n+nf}{integral}\PY{p}{(}\PY{n}{default}\PY{p}{,} \PY{n}{companies}\PY{p}{,} \PY{n}{t}\PY{p}{)}\PY{p}{:}
    \PY{k}{return} \PY{n}{quad}\PY{p}{(}\PY{n}{func}\PY{p}{,} \PY{l+m+mi}{0}\PY{p}{,} \PY{n}{default}\PY{p}{[}\PY{n}{companies}\PY{p}{[}\PY{l+m+mi}{0}\PY{p}{]}\PY{p}{]}\PY{p}{[}\PY{n}{t}\PY{p}{]}\PY{p}{,} 

\PY{n}{args}\PY{o}{=}\PY{p}{(}\PY{n}{default}\PY{p}{,} \PY{n}{companies}\PY{p}{,} \PY{n}{t}\PY{p}{)}\PY{p}{)}\PY{p}{[}\PY{l+m+mi}{0}\PY{p}{]}
\PY{k}{for} \PY{n}{companies} \PY{o+ow}{in} \PY{p}{[}\PY{p}{(}\PY{l+s+s2}{\PYZdq{}}\PY{l+s+s2}{A}\PY{l+s+s2}{\PYZdq{}}\PY{p}{,} \PY{l+s+s2}{\PYZdq{}}\PY{l+s+s2}{B}\PY{l+s+s2}{\PYZdq{}}\PY{p}{,} \PY{l+s+s2}{\PYZdq{}}\PY{l+s+s2}{C}\PY{l+s+s2}{\PYZdq{}}\PY{p}{)}\PY{p}{,} \PY{p}{(}\PY{l+s+s2}{\PYZdq{}}\PY{l+s+s2}{B}\PY{l+s+s2}{\PYZdq{}}\PY{p}{,} \PY{l+s+s2}{\PYZdq{}}\PY{l+s+s2}{A}\PY{l+s+s2}{\PYZdq{}}\PY{p}{,} \PY{l+s+s2}{\PYZdq{}}\PY{l+s+s2}{C}\PY{l+s+s2}{\PYZdq{}}\PY{p}{)}\PY{p}{,} \PY{p}{(}\PY{l+s+s2}{\PYZdq{}}\PY{l+s+s2}{C}\PY{l+s+s2}{\PYZdq{}}\PY{p}{,} \PY{l+s+s2}{\PYZdq{}}\PY{l+s+s2}{A}\PY{l+s+s2}{\PYZdq{}}\PY{p}{,} \PY{l+s+s2}{\PYZdq{}}\PY{l+s+s2}{B}\PY{l+s+s2}{\PYZdq{}}\PY{p}{)}\PY{p}{]}\PY{p}{:}
    \PY{n}{prob} \PY{o}{=} \PY{l+m+mi}{0}
    \PY{k}{for} \PY{n}{t} \PY{o+ow}{in} \PY{n+nb}{range}\PY{p}{(}\PY{l+m+mi}{1}\PY{p}{,} \PY{l+m+mi}{6}\PY{p}{)}\PY{p}{:}
    \PY{n}{prob} \PY{o}{=} \PY{n}{integral}\PY{p}{(}\PY{n}{default\PYZus{}rates}\PY{p}{,} \PY{n}{companies}\PY{p}{,} \PY{n}{t}\PY{p}{)}
    \PY{n+nb}{print} \PY{p}{(}\PY{l+s+s2}{\PYZdq{}}\PY{l+s+s2}{P(1st def) at time (}\PY{l+s+si}{\PYZob{}\PYZcb{}}\PY{l+s+s2}{) for company }\PY{l+s+si}{\PYZob{}\PYZcb{}}\PY{l+s+s2}{: }\PY{l+s+si}{\PYZob{}:.5f\PYZcb{}}\PY{l+s+s2}{\PYZdq{}}\PY{o}{.}\PY{n}{format}\PY{p}{(}\PY{n}{t}\PY{p}{,} 
    \PY{n}{companies}\PY{p}{[}\PY{l+m+mi}{0}\PY{p}{]}\PY{p}{,} \PY{n}{prob}\PY{p}{)}\PY{p}{)}

First to default prob at time (1) for company A: 0.02131
First to default prob at time (2) for company A: 0.04301
First to default prob at time (3) for company A: 0.06460
First to default prob at time (4) for company A: 0.08573
First to default prob at time (5) for company A: 0.10606
First to default prob at time (1) for company B: 0.03080
First to default prob at time (2) for company B: 0.06160
...
\end{Verbatim}
\end{tcolorbox}

\subsubsection{Correlated Defaults}\label{correlated-defaults}
When the default probabilities of the companies are correlated the copula approach can be used like in the example shown in Section~\ref{generate-correlated-distributions}.

Suppose to have to estimate the default probabilities for the next 5 years for 6 companies. 
The copula default correlation between each company is 0.2 and the cumulative probability of default during the next 1,2,3,4 5 years is 1\%, 3\%, 6\%, 10\%, 13\% respectively for each company.

When a Gaussian copula is used in order to simulate the defaults we need to sample from a multivariate normal distribution a vector $\mathbf{x}$, transform then each $x_i$ into the corresponding default probability $p_i$.

Let's check the 3th-to-default probabilities for each year.
\begin{tcolorbox}[breakable, size=fbox, boxrule=1pt, pad at break*=1mm,colback=cellbackground, colframe=cellborder]
\begin{Verbatim}[commandchars=\\\{\}]
\PY{k+kn}{from} \PY{n+nn}{scipy}\PY{n+nn}{.}\PY{n+nn}{stats} \PY{k}{import} \PY{n}{multivariate\PYZus{}normal}
	
\PY{n}{p\PYZus{}default} \PY{o}{=} \PY{p}{[}\PY{l+m+mi}{0}\PY{p}{,} \PY{l+m+mf}{0.01}\PY{p}{,} \PY{l+m+mf}{0.03}\PY{p}{,} \PY{l+m+mf}{0.06}\PY{p}{,} \PY{l+m+mf}{0.10}\PY{p}{,} \PY{l+m+mf}{0.13}\PY{p}{]}
	
\PY{n}{mvnorm} \PY{o}{=} \PY{n}{multivariate\PYZus{}normal}\PY{p}{(}\PY{n}{mean}\PY{o}{=}\PY{p}{[}\PY{l+m+mi}{0}\PY{p}{]}\PY{o}{*}\PY{l+m+mi}{6}\PY{p}{,}
                             \PY{n}{cov} \PY{o}{=} \PY{p}{[}\PY{p}{[}\PY{l+m+mi}{1}\PY{p}{,} \PY{l+m+mf}{0.2}\PY{p}{,} \PY{l+m+mf}{0.2}\PY{p}{,} \PY{l+m+mf}{0.2}\PY{p}{,} \PY{l+m+mf}{0.2}\PY{p}{,} \PY{l+m+mf}{0.2}\PY{p}{]}\PY{p}{,}
                                    \PY{p}{[}\PY{l+m+mf}{0.2}\PY{p}{,} \PY{l+m+mi}{1}\PY{p}{,} \PY{l+m+mf}{0.2}\PY{p}{,} \PY{l+m+mf}{0.2}\PY{p}{,} \PY{l+m+mf}{0.2}\PY{p}{,} \PY{l+m+mf}{0.2}\PY{p}{]}\PY{p}{,}
                                    \PY{p}{[}\PY{l+m+mf}{0.2}\PY{p}{,} \PY{l+m+mf}{0.2}\PY{p}{,} \PY{l+m+mi}{1}\PY{p}{,} \PY{l+m+mf}{0.2}\PY{p}{,} \PY{l+m+mf}{0.2}\PY{p}{,} \PY{l+m+mf}{0.2}\PY{p}{]}\PY{p}{,}
                                    \PY{p}{[}\PY{l+m+mf}{0.2}\PY{p}{,} \PY{l+m+mf}{0.2}\PY{p}{,} \PY{l+m+mf}{0.2}\PY{p}{,} \PY{l+m+mi}{1}\PY{p}{,} \PY{l+m+mf}{0.2}\PY{p}{,} \PY{l+m+mf}{0.2}\PY{p}{]}\PY{p}{,}
                                    \PY{p}{[}\PY{l+m+mf}{0.2}\PY{p}{,} \PY{l+m+mf}{0.2}\PY{p}{,} \PY{l+m+mf}{0.2}\PY{p}{,} \PY{l+m+mf}{0.2}\PY{p}{,} \PY{l+m+mi}{1}\PY{p}{,} \PY{l+m+mf}{0.2}\PY{p}{]}\PY{p}{,}
                                    \PY{p}{[}\PY{l+m+mf}{0.2}\PY{p}{,} \PY{l+m+mf}{0.2}\PY{p}{,} \PY{l+m+mf}{0.2}\PY{p}{,} \PY{l+m+mf}{0.2}\PY{p}{,} \PY{l+m+mf}{0.2}\PY{p}{,} \PY{l+m+mi}{1}\PY{p}{]}\PY{p}{]}\PY{p}{)}
	
\PY{n}{trials} \PY{o}{=} \PY{l+m+mi}{100000}
\PY{n}{result} \PY{o}{=} \PY{p}{[}\PY{l+m+mf}{0.}\PY{p}{,} \PY{l+m+mf}{0.}\PY{p}{,} \PY{l+m+mf}{0.}\PY{p}{,} \PY{l+m+mf}{0.}\PY{p}{,} \PY{l+m+mf}{0.}\PY{p}{,} \PY{l+m+mf}{0.}\PY{p}{]}
\PY{n}{x} \PY{o}{=} \PY{n}{mvnorm}\PY{o}{.}\PY{n}{rvs}\PY{p}{(}\PY{n}{size}\PY{o}{=}\PY{n}{trials}\PY{p}{)}
	
\PY{k}{for} \PY{n}{n} \PY{o+ow}{in} \PY{n+nb}{range}\PY{p}{(}\PY{n+nb}{len}\PY{p}{(}\PY{n}{x}\PY{p}{)}\PY{p}{)}\PY{p}{:}
    \PY{n}{p} \PY{o}{=} \PY{n+nb}{sorted}\PY{p}{(}\PY{n}{norm}\PY{o}{.}\PY{n}{cdf}\PY{p}{(}\PY{n}{x}\PY{p}{[}\PY{n}{n}\PY{p}{]}\PY{p}{)}\PY{p}{)}
    \PY{k}{for} \PY{n}{i} \PY{o+ow}{in} \PY{n+nb}{range}\PY{p}{(}\PY{l+m+mi}{1}\PY{p}{,} \PY{n+nb}{len}\PY{p}{(}\PY{n}{p\PYZus{}default}\PY{p}{)}\PY{p}{)}\PY{p}{:}
        \PY{k}{if} \PY{n}{p\PYZus{}default}\PY{p}{[}\PY{n}{i}\PY{o}{\PYZhy{}}\PY{l+m+mi}{1}\PY{p}{]} \PY{o}{\PYZlt{}}\PY{o}{=} \PY{n}{p}\PY{p}{[}\PY{l+m+mi}{2}\PY{p}{]} \PY{o}{\PYZlt{}}\PY{o}{=} \PY{n}{p\PYZus{}default}\PY{p}{[}\PY{n}{i}\PY{p}{]}\PY{p}{:}
            \PY{n}{result}\PY{p}{[}\PY{n}{i}\PY{p}{]} \PY{o}{+}\PY{o}{=} \PY{l+m+mi}{1}
	
\PY{n+nb}{print} \PY{p}{(}\PY{l+s+s2}{\PYZdq{}}\PY{l+s+s2}{3rd\PYZhy{}to\PYZhy{}default probabilies}\PY{l+s+s2}{\PYZdq{}}\PY{p}{)}
\PY{k}{for} \PY{n}{i} \PY{o+ow}{in} \PY{n+nb}{range}\PY{p}{(}\PY{n+nb}{len}\PY{p}{(}\PY{n}{p\PYZus{}default}\PY{p}{)}\PY{p}{)}\PY{p}{:}
    \PY{n+nb}{print} \PY{p}{(}\PY{l+s+s2}{\PYZdq{}}\PY{l+s+si}{\PYZob{}\PYZcb{}}\PY{l+s+s2}{: }\PY{l+s+si}{\PYZob{}:.4f\PYZcb{}}\PY{l+s+s2}{\PYZdq{}}\PY{o}{.}\PY{n}{format}\PY{p}{(}\PY{n}{i}\PY{p}{,} \PY{n}{result}\PY{p}{[}\PY{n}{i}\PY{p}{]}\PY{o}{/}\PY{n}{trials}\PY{p}{)}\PY{p}{)}

3rd-to-default probabilies
0: 0.0000
1: 0.0003
2: 0.0033
3: 0.0109
4: 0.0250
5: 0.0267
\end{Verbatim}
\end{tcolorbox}


\section{Basket CDS Valuation under Gaussian Copula Model}\label{basket-cds-valuation-under-market-standard-model}
We now present some numerical results for an nth-to-default basket.
We assume that the principals and expected recovery rates are the same
for all underlying reference assets. The valuation procedure is similar
to that for a regular CDS where there is only one reference entity.

In a regular CDS indeed the valuation is based on the probability that a
default occurred between times \(t1\) and \(t2\). Here instead the
valuation will be based on the probability that the nth default was
between times \(t1\) and \(t2\).
The buyer of protection makes quarterly payments at a
specified rate until the $n^{th}$ default occurs or the end life
of the contract is reached.

In the event of the $n^{th}$ default occurring, the seller pays
\(F\cdot(1-R)\). The contract can be valued in a similar way as done for the CDS.

So consider a 5-year nth-to-default basket of ten
reference entities in the situation where the copula correlation is 0.3
and the expected recovery rate, \(R\), is \(40\%\). The term structure
of interest rates is assumed to be flat at 5\%. The default
probabilities for the ten entities are generated by Poisson processes
with constant default intensities (hazard rates), \(\lambda_i\),
\((1 \le i \le 10)\) so that

\[ DP(t) = 1 - e^{-\lambda t} \]

\begin{tcolorbox}[breakable, size=fbox, boxrule=1pt, pad at break*=1mm,colback=cellbackground, colframe=cellborder]
\begin{Verbatim}[commandchars=\\\{\}]
\PY{k+kn}{from} \PY{n+nn}{finmarkets} \PY{k}{import} \PY{n}{DiscountCurve}\PY{p}{,} \PY{n}{CreditCurve}\PY{p}{,} \PY{n}{CreditDefaultSwap}
\PY{k+kn}{from} \PY{n+nn}{finmarkets} \PY{k}{import} \PY{n}{GaussianQuadrature}
\PY{k+kn}{from} \PY{n+nn}{datetime} \PY{k}{import} \PY{n}{date}
\PY{k+kn}{from} \PY{n+nn}{dateutil}\PY{n+nn}{.}\PY{n+nn}{relativedelta} \PY{k}{import} \PY{n}{relativedelta}
\PY{k+kn}{from} \PY{n+nn}{scipy}\PY{n+nn}{.}\PY{n+nn}{stats} \PY{k}{import} \PY{n}{norm}\PY{p}{,} \PY{n}{binom}
\PY{k+kn}{from} \PY{n+nn}{math} \PY{k}{import} \PY{n}{sqrt}\PY{p}{,} \PY{n}{exp}
	
\PY{n}{n\PYZus{}cds} \PY{o}{=} \PY{l+m+mi}{10}
\PY{n}{rho} \PY{o}{=} \PY{l+m+mf}{0.3}
\PY{n}{l} \PY{o}{=} \PY{l+m+mf}{0.01}
\PY{n}{pillar\PYZus{}dates} \PY{o}{=} \PY{p}{[}\PY{p}{]}
\PY{n}{df} \PY{o}{=} \PY{p}{[}\PY{p}{]}
\PY{n}{observation\PYZus{}date} \PY{o}{=} \PY{n}{date}\PY{o}{.}\PY{n}{today}\PY{p}{(}\PY{p}{)}
	
\PY{k}{for} \PY{n}{i} \PY{o+ow}{in} \PY{n+nb}{range}\PY{p}{(}\PY{l+m+mi}{6}\PY{p}{)}\PY{p}{:}
\PY{n}{pillar\PYZus{}dates}\PY{o}{.}\PY{n}{append}\PY{p}{(}\PY{n}{observation\PYZus{}date} \PY{o}{+} \PY{n}{relativedelta}\PY{p}{(}\PY{n}{years}\PY{o}{=}\PY{n}{i}\PY{p}{)}\PY{p}{)}
\PY{n}{df}\PY{o}{.}\PY{n}{append}\PY{p}{(}\PY{l+m+mi}{1}\PY{o}{/}\PY{p}{(}\PY{l+m+mi}{1}\PY{o}{+}\PY{l+m+mf}{0.05}\PY{o}{*}\PY{n}{i}\PY{p}{)}\PY{p}{)}
\PY{n}{dc} \PY{o}{=} \PY{n}{DiscountCurve}\PY{p}{(}\PY{n}{observation\PYZus{}date}\PY{p}{,} \PY{n}{pillar\PYZus{}dates}\PY{p}{,} \PY{n}{df}\PY{p}{)}
	
\PY{n}{gq} \PY{o}{=} \PY{n}{GaussianQuadrature}\PY{p}{(}\PY{p}{)}
\PY{n}{values}\PY{p}{,} \PY{n}{weights} \PY{o}{=} \PY{n}{gq}\PY{o}{.}\PY{n}{M}\PY{p}{(}\PY{l+m+mi}{60}\PY{p}{)}
	
\PY{n}{Q} \PY{o}{=} \PY{p}{[}\PY{l+m+mi}{1}\PY{o}{\PYZhy{}}\PY{n}{exp}\PY{p}{(}\PY{o}{\PYZhy{}}\PY{p}{(}\PY{n}{l}\PY{o}{*}\PY{n}{t}\PY{p}{)}\PY{p}{)} \PY{k}{for} \PY{n}{t} \PY{o+ow}{in} \PY{n+nb}{range}\PY{p}{(}\PY{l+m+mi}{6}\PY{p}{)}\PY{p}{]}
\PY{n}{cds} \PY{o}{=} \PY{n}{CreditDefaultSwap}\PY{p}{(}\PY{l+m+mi}{1}\PY{p}{,} \PY{n}{observation\PYZus{}date}\PY{p}{,} \PY{l+m+mf}{0.01}\PY{p}{,} \PY{l+m+mi}{5}\PY{p}{)}
	
\PY{n}{ndefault} \PY{o}{=} \PY{l+m+mi}{3}
\PY{n}{S} \PY{o}{=} \PY{p}{[}\PY{p}{]}
\PY{k}{for} \PY{n}{j} \PY{o+ow}{in} \PY{n+nb}{range}\PY{p}{(}\PY{n+nb}{len}\PY{p}{(}\PY{n}{values}\PY{p}{)}\PY{p}{)}\PY{p}{:}
    \PY{n}{temp} \PY{o}{=} \PY{p}{[}\PY{p}{]}
    \PY{k}{for} \PY{n}{i} \PY{o+ow}{in} \PY{n+nb}{range}\PY{p}{(}\PY{l+m+mi}{6}\PY{p}{)}\PY{p}{:}
        \PY{n}{P} \PY{o}{=} \PY{n}{norm}\PY{o}{.}\PY{n}{cdf}\PY{p}{(}\PY{p}{(}\PY{n}{norm}\PY{o}{.}\PY{n}{ppf}\PY{p}{(}\PY{n}{Q}\PY{p}{[}\PY{n}{i}\PY{p}{]}\PY{p}{)} \PY{o}{\PYZhy{}} \PY{n}{sqrt}\PY{p}{(}\PY{n}{rho}\PY{p}{)}\PY{o}{*}\PY{n}{values}\PY{p}{[}\PY{n}{j}\PY{p}{]}\PY{p}{)}\PY{o}{/}
        \PY{p}{(}\PY{n}{sqrt}\PY{p}{(}\PY{l+m+mi}{1}\PY{o}{\PYZhy{}}\PY{n}{rho}\PY{p}{)}\PY{p}{)}\PY{p}{)}
        \PY{n}{b} \PY{o}{=} \PY{n}{binom}\PY{p}{(}\PY{n}{n\PYZus{}cds}\PY{p}{,} \PY{n}{P}\PY{p}{)}
        \PY{n}{temp}\PY{o}{.}\PY{n}{append}\PY{p}{(}\PY{l+m+mi}{1} \PY{o}{\PYZhy{}} \PY{p}{(}\PY{n}{b}\PY{o}{.}\PY{n}{cdf}\PY{p}{(}\PY{n}{n\PYZus{}cds}\PY{p}{)}\PY{o}{\PYZhy{}}\PY{n}{b}\PY{o}{.}\PY{n}{cdf}\PY{p}{(}\PY{n}{ndefault}\PY{o}{\PYZhy{}}\PY{l+m+mi}{1}\PY{p}{)}\PY{p}{)}\PY{p}{)}
    \PY{n}{S}\PY{o}{.}\PY{n}{append}\PY{p}{(}\PY{n}{temp}\PY{p}{)}
	
\PY{n}{s} \PY{o}{=} \PY{l+m+mi}{0}
\PY{k}{for} \PY{n}{j} \PY{o+ow}{in} \PY{n+nb}{range}\PY{p}{(}\PY{n+nb}{len}\PY{p}{(}\PY{n}{values}\PY{p}{)}\PY{p}{)}\PY{p}{:}
    \PY{n}{s} \PY{o}{+}\PY{o}{=} \PY{n}{weights}\PY{p}{[}\PY{n}{j}\PY{p}{]} \PY{o}{*} \PY{n}{cds}\PY{o}{.}\PY{n}{breakevenRate}\PY{p}{(}\PY{n}{dc}\PY{p}{,} \PY{n}{CreditCurve}\PY{p}{(}\PY{n}{pillar\PYZus{}dates}\PY{p}{,} \PY{n}{S}\PY{p}{[}\PY{n}{j}\PY{p}{]}\PY{p}{)}\PY{p}{)}
\PY{n+nb}{print} \PY{p}{(}\PY{n}{s}\PY{p}{)}

0.0017795634956118353
\end{Verbatim}
\end{tcolorbox}














\section{Collateralized Debt Obligation}\label{collateralized-debt-obligation}

A Collateralized Debt Obligation (CDO) is a credit derivative1
that creates fixed income securities with widely different risk characteristics from a pool of risky assets. 
The
coupon and principal payments of these securities are linked to the performance of the
underlying pool. These fixed income securities are known as tranches and divided into
senior, mezzanine and subordinated/equity tranches. 

Each of these tranches has a different level of seniority relative to the others in the sense that a senior tranche has coupon
and principal payment priority over a mezzanine tranche, while a mezzanine tranche has
coupon and principal payment priority over an equity tranche. 


It is important to note
that a CDO only redistributes the total risk associated with the underlying pool of assets
to the priority ordered tranches. It neither reduces nor increases the total risk associated
with the pool.

A CDO is called a synthetic CDO if the risky assets in the underlying pool are credit default swaps (CDS). 
Besides synthetic CDOs,
there are other types of CDOs. Depending on the nature of the risky assets in the
underlying pool, a CDO may be called a collateralized loan obligation or a collateralized
bond obligation if it holds only loans or bonds, respectively. 

the transaction is called a balance sheet CDO.
A CDO can be structured as either a cash flow or a synthetic transaction, or a hybrid
of both. In a cash flow transaction, the CDO is backed by a pool of cash assets that are
truly owned by the CDO sponsor. A synthetic CDO makes use of CDS’ to transfer the
credit risk of a pool of reference entities to tranche investors. 
The main difference between
a cash flow CDO and a synthetic CDO is that in the latter no transfer of securities takes
place. 

\subsection{CDO Expected Losses}


\subsection{CDO Valuation}
The structure of a typical synthetic CDO can be explained through the following example,
which is illustrated in Figure 1.1. This synthetic CDO is based on a pool of 125 CDS’. The
notional value of each CDS is US$8 million. The total notional value of the underlying
pool is US$1 billion. The CDO has four priority ordered tranches: (i) an equity tranche
with an initial notional value of US$30 million and a credit spread of 2000 basis points
(bps) per annum; (ii) a mezzanine tranche with an initial notional value of US$50 million
and a credit spread of 800 bps per annum; (iii) a senior tranche with an initial notional
value of US$120 million and a credit spread of 100 bps per annum; and (iv) a super
senior tranche with an initial notional value of US$800 million and a credit spread of 20
bps per annum. The equity tranche is the riskiest one; the mezzanine tranche bears a
medium risk; the senior tranche is less risky and the super senior tranche has the lowest
risk. The maturity of the CDO, thus the maturity of each tranche, is five years from now
and the premiums are paid quarterly.
During the life of the CDO, each tranche may receive premiums quarterly from the
CDO issuer. The premium that a tranche investor receives for a specified premium
payment period is proportional to the remaining notional value of the specified tranche
at the end of the premium payment period. Let us consider one scenario. Suppose there
is no loss in the first quarter of the CDO life. Then the equity tranche investor receives
30 ∗ 20% ∗ 0.25 =US$1.5 million from the CDO issuer; the mezzanine tranche investor
receives 50 ∗ 8% ∗ 0.25 =US$1 million, and so on. Suppose further that, in the second
quarter, the underlying pool suffers a US$6 million loss. Then the equity tranche investor
absorbs this US$6 million loss. The premium that the equity tranche investor receives
for this period is (30 − 6) ∗ 20% ∗ 0.25 =US$1.2 million. In total, the equity tranche
investor pays the CDO issuer US$(6-1.2)=US$4.8 million. The notional value of the
equity tranche for the next premium payment period is reduced to US$(30-6)=US$24
Chapter 1. Introduction 5
million. The mezzanine tranche investor still receives 50 ∗ 8% ∗ 0.25 =US$1 million from
the CDO issuer, and so on. If the equity tranche’s notional value has been reduced to
zero before the maturity of the CDO, the mezzanine tranche investor would then be the
next one to suffer losses. The pool would need to suffer a loss of US$200 million before
the super senior tranche investor would suffer any loss.




























%A collateralized debt obligation (CDO) is a type of structured asset-backed security (ABS). 
%A CDO can be thought of as a promise to pay investors in a prescribed sequence, based on the cash flow the CDO collects from the pool of bonds or other assets it owns. Distinctively, CDO credit risk is typically assessed based on a probability of default (PD) derived from ratings on those bonds or assets.

%The CDO is "sliced" into "tranches", which "catch" the cash flow of interest and principal payments in sequence based on seniority.
%In the literature, tranches of a CDO are classified as
%subordinate/equity tranche, mezzanine tranches, and senior tranches
%according to their levels.
%
%If some loans default and the cash collected by the CDO is insufficient to pay all of its investors, those in the lowest, most "junior" tranches suffer losses first. The last to lose payment from default are the safest, most senior tranches. Consequently, coupon payments (and interest rates) vary by tranche with the safest/most senior tranches receiving the lowest rates and the lowest tranches receiving the highest rates to compensate for higher default risk.
%
%Each tranche has an attachment percentage and a detachment
%percentage. When the cumulative percentage loss of the portfolio reaches
%the attachment percentage, investors in the tranche start to lose their
%principal, and when the cumulative percentage loss of principal reaches
%the detachment percentage, the investors in the tranche lose all their
%principal and no further loss can occur to them.
%
%A common analogy compares the cash flow from the CDO's portfolio of securities (say mortgage payments from mortgage-backed bonds) to water flowing into cups of the investors where senior tranches were filled first and overflowing cash flowed to junior tranches, then equity tranches. If a large portion of the mortgages enter default, there is insufficient cash flow to fill all these cups and equity tranche investors face the losses first.



%\section{CDO Valuation}
%Suppose that the payment dates of a CDO tranche are at times $d_i$. Define $\mathbb{E}_j$ the expected 
%tranche value at time $d$ and $D(d)$ the discount factor. Suppose also that the spread
%on a particular tranche (i.e. the number of basis point paid for protection on the remaining tranche principal) is $S$. 
%
%The present value of the expected regular spread payments on the CDO is given by
%\begin{equation}
%\mathrm{NPV}_{\mathrm{spread}} = S\cdot \sum_{j=1}^{m}(d_j - d_{j-1})\mathbb{E}_{j}D(d_j)
%\label{eq:A}
%\end{equation}
%The expected loss between times $d_{j-1}$ and $d_j$ is $\mathbb{E}_{j-1}-\mathbb{E}_j$. For simplicity assume
%the loss occurs only at the midpoint of the time interval, so the present value of the expected payoffs on the CDO tranche is
%\begin{equation}
%\mathrm{NPV}_{\mathrm{payoff}}=\sum_{j=1}^{m}(\mathbb{E}_{j-1}-\mathbb{E}_j)D\left(\frac{d_{j-1}+d_j}{2}\right)
%\label{eq:C}
%\end{equation}
%The accrual payment due to the losses is finally given by
%\begin{equation}
%\mathrm{NPV}_{\mathrm{accrual}} = S\cdot\sum_{j=1}^{m}\frac{1}{2}(d_j - d_{j-1})(\mathbb{E}_{j-1}-\mathbb{E}_j)D(\frac{d_{j-1}+d_j}{2})
%\label{eq:B}
%\end{equation}
%
%The value of the tranche, valued from the point of view of the protection buyer is $C-sA-sB$. The breakeven spread 
%on the tranche occurs when the present value of the payments equals the present value of the payoffs so
%
%\[ s = \cfrac{C}{A+B}\]
%
%Suppose that the tranche under consideration covers losses on the portfolio between $\alpha_L$ and $\alpha_H$ and
%define
%
%\[n_L = \cfrac{\alpha_L n}{1-R}\qquad \mathrm{and}\qquad n_H = \cfrac{\alpha_H n}{1-R}\]
%where $R$ is the recovery rate. Finally define $m(x)$ as the smallest integer greater than $x$.
%By definition the tranche principal stays $N$ while the number of defaults $k$ is less than $m(n_L)$, it 
%is zero when the number of default is greater or equal to $m(n_H)$, otherwise is
%
%\[\cfrac{\alpha_H -k(1-R)/n}{\alpha_H - \alpha_L}\]
%
%The expected tranche principal at time $\tau_j$ conditional of the value of the factor $M$ is
%\begin{equation}
%\mathbb{E}_j(M) = \sum_{k=0}^{m(n_L)-1}\mathbb{P}(k, \tau_j|M) + \sum_{k=m(n_L)}^{m(n_H)-1}\mathbb{P}(k, \tau_j|M) \cfrac{\alpha_H -k(1-R)/n}{\alpha_H - \alpha_L}
%\label{eq:E}
%\end{equation}
%
%To compute the breakeven spread it is necessary to substitute Eq.~\ref{eq:E} into Eq.~\ref{eq:A},~\ref{eq:B} and~\ref{eq:C}
%and we need to integrate the result over the variable $M$ (remember that has a standard normal distribution). 
%The integration is quite complicated and is best accomplished with a technique called \emph{Gaussian quadrature} which
%exploits the approximation
%
%\[\int_{-\infty}^{\infty}{\cfrac{1}{\sqrt{2}}e^{-M^{2}/2}g(M)dM} \approx \sum_{k=1}^{k=L}w_k g(F_k)\]
%as $L$ increases, accuracy increases.


\section{Complex Correlation Structures and the Financial
	Crisis}\label{complex-correlation-structures-and-the-financial-crisis}

In the derivation of the Gaussian Copula Model we have used the normal distributions. However, we could have used other and more complex
copulas as well. For example we might want to assume the correlation is
non-symmetric which is useful in finance when correlations become
very strong during market crashes and returns very negative.

In fact, Gaussian copulas are said to have played a key role in the 2008
financial crisis as tail-correlations were severely underestimated.
Consider a set of mortgages in a CDO (a particular kind of contract that will be introduced in the next Chapter): they are clearly correlated, since if one mortgage fails,
the likelihood that another failing is increased. In the early 2000s,
the banks only knew how to model the marginals of the default rates. Then an
(in)famous paper by Li suggested to use Gaussian copulas to model the
correlations between those marginals. Rating agencies relied on this model so heavily, that severely underestimated the risk and gave false ratings\ldots

If you are interested in the argument read
\href{http://samueldwatts.com/wp-content/uploads/2016/08/Watts-Gaussian-Copula_Financial_Crisis.pdf}{this paper}
for an excellent description of Gaussian copulas and the Financial
Crisis, which argues that different copula choices would not have made a
difference but instead the assumed correlation was way too low.
